---
title: "Quality Control of SHARK Data"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quality Control of SHARK Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Quality Control Overview

The SHARK4R package provides a set of functions to perform quality control on SHARK occurrence data. These functions help identify missing or invalid values, spatial errors, and statistical outliers.


#### Checking Required Fields

- `check_datatype()` verifies that all global SHARK required fields are present in an occurrence table and checks for missing values. It returns a data frame listing any errors found.  
- `check_fields()` validates datatype-specific fields (defined in the parameters) or values. Required and recommended fields are defined using the parameter `field_definitions` (e.g. `SHARK4R:::.threshold_values` as default). 

#### Geospatial Checks

- `plot_map_leaflet()` creates an interactive Leaflet map to visualize your data points.  
- `check_onland()` uses the `lookup_xy()` OBIS web service to detect points located on land. You can also use alternative shapefiles if needed.  
- `positions_are_near_land()` and `which_basin()` from the [iRfcb package](https://CRAN.R-project.org/package=iRfcb) can also be used for geospatial quality control.

#### Depth Checks

`check_depth()` performs several checks on depth values, including:

- Missing depth column (warning)  
- Empty depth values (warning)  
- Non-numeric values (error)  
- Depth exceeding bathymetry layer with margin using the `lookup_xy()` OBIS web service (error)  
- Negative depth offshore with shore margin (error)  
- Minimum depth greater than maximum depth (error)

#### Outlier Detection

`check_outliers()` identifies parameter values that exceed predefined thresholds. Thresholds are provided in a data frame (e.g., `SHARK4R:::.threshold_values` or retrieved from `get_shark_statistics()`), which must include columns for `parameter`, `datatype`, and at least one numeric threshold column (e.g., `extreme_upper`).  

Key points:

- Only rows in `data` matching both the specified `parameter` and `datatype` are considered.  
- The threshold column to check (e.g., `extreme_upper`) can be specified via `threshold_col`.  
- Values exceeding the threshold are flagged as outliers and returned in an interactive table using `DT::datatable()`.  
- If all values are within the threshold, a message is printed to indicate that no outliers were detected.  

This approach allows flexible, parameter-specific QC based on SHARK-defined thresholds rather than generic statistical deviation rules. The `scatterplot()` function can be used to visualize the data.

#### Interactive Quality Control with Shiny

Instead of running individual QC functions manually as demonstrated below, you can use a **Shiny app** that provides an interactive interface for checking SHARK data.

The app is part of the SHARK4R support files and can be downloaded and initialized with:

```{r, eval=FALSE}
# Download support files into current working directory
check_setup()

# Run Shiny app
shiny::runApp("products/")

# Or download and run the QC Shiny app directly
check_setup(run_app = TRUE)
```

The Shiny app includes tools for running the same QC checks described above, but with a point-and-click interface. This is particularly useful for exploring datasets interactively and for users less familiar with R scripting.

---

## Using the QC-functions

### Installation

You can install the latest version of the package from GitHub using the `remotes` package:
```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("sharksmhi/SHARK4R",
                        ref = remotes::github_release(),
                        dependencies = TRUE)
```

Once installed, load the package along with `dplyr` for data manipulation::
```{r, include=FALSE}
suppressPackageStartupMessages({
  library(SHARK4R)
  library(dplyr)
})
```

```{r, eval=FALSE}
library(SHARK4R)
library(dplyr)
```

### Retrieve Data Table

You can fetch SHARK data using the same filtering options as the SHARK web interface. Available options can be explored using `get_shark_options()`. For more information, see the the [Retrieve Data From SHARK](https://sharksmhi.github.io/SHARK4R/articles/retrieve_shark_data.html) tutorial.

```{r}
# Retrieve available options from SHARK
shark_options <- get_shark_options()
```

You can filter datasets to focus on specific parameters, such as "Chlorophyll":

```{r}
# Find datasets containing "Chlorophyll"
chlorophyll_datasets <- shark_options$datasets[grepl("Chlorophyll", 
                                                   shark_options$datasets)]

# Select first matching dataset
selected_dataset <- chlorophyll_datasets[1]

# Print selected dataset name
print(selected_dataset)
```

Download the dataset and store it as a data frame:

```{r}
# Download dataset and return as data frame
chlorophyll_data <- get_shark_datasets(selected_dataset,
                                       save_dir = tempdir(),
                                       return_df = TRUE,
                                       verbose = FALSE)

# Print data
tibble(chlorophyll_data)
```

### Checking Fields

Once the data is loaded, it is important to validate that all required fields are present and correctly formatted. `check_fields()` performs this check for datatype-specific fields:

```{r}
# Check datatype-specific fields for Chlorophyll data
check_fields(data = chlorophyll_data, 
             datatype = "Chlorophyll")
```

### Station Matching

SHARK data often contains station names that need to be verified against the official SHARK station registry. This ensures correct geographic and metadata matching:

```{r}
# Match station names against SHARK station registry
station_match <- match_station(chlorophyll_data$station_name)

# Show first rows of matches
head(station_match)
```

### Points on Land

Sometimes, geographic coordinates are misreported and fall on land instead of the ocean. Use `check_onland()` to identify such points:

```{r}
# Identify rows with coordinates falling on land
n_rows_on_land <- check_onland(chlorophyll_data)

# Count number of flagged rows
nrow(n_rows_on_land)
```

### Checking Depth

Depth values are critical for aquatic measurements. `check_depth()` verifies that reported depths are plausible and consistent:

```{r}
# Check min/max depth
check_depth(data = chlorophyll_data)

# Check if reported water depth column is deeper than bathymetry
check_depth(data = chlorophyll_data, "water_depth_m")
```

This helps detect missing values, negative depths, or depths exceeding local bathymetry.

### Code Validation

Project or platform codes must follow SHARK conventions. `check_codes()` helps validate these codes:

```{r}
# Validate project codes
check_codes(chlorophyll_data)

# Validate ship/platform codes
check_codes(data = chlorophyll_data, 
            field = "platform_code", 
            code_type = "SHIPC", 
            match_column = "Code")
```

### Retrieve Statistics

Some QC checks, like outlier detection, rely on reference statistics. `get_shark_statistics()` provides thresholds for different parameters and datatypes:

```{r}
# Retrieve SHARK statistics for Chlorophyll (2020–2024)
shark_statistics <- get_shark_statistics(datatype = "Chlorophyll",
                                         fromYear = 2020,
                                         toYear = 2024,
                                         verbose = FALSE)

# Print statistics
tibble(shark_statistics)
```

These statistics can be used to identify extreme or unusual values in your data.

### Outlier Detection

`check_outliers()` flags values that exceed thresholds defined in SHARK statistics. This is useful for identifying unusual measurements that may indicate errors or extreme events:

```{r}
# Detect Chlorophyll-a outliers using the 99th percentile (P99)
check_outliers(data = chlorophyll_data,
               parameter = "Chlorophyll-a",
               datatype = "Chlorophyll",
               threshold_col = "P99",
               thresholds = shark_statistics)
```

### Data Visualization

Visualizing data geographically or as scatterplots is useful for spotting anomalies quickly.

```{r}
# Visualize data on an interactive map
plot_map_leaflet(chlorophyll_data)
```

```{r}
# Create an interactive scatterplot of parameter values
scatterplot(chlorophyll_data)
```

Maps help verify the location of points, while scatterplots highlight unusual measurements across stations or time.

---

## Recommended Workflow

For a comprehensive quality control of SHARK data, we recommend the following order of steps:

1. **Check Required Fields**  
   Validate all mandatory fields using `check_datatype()` and datatype-specific fields with `check_fields()`.

2. **Validate Codes**  
   Check project and platform codes with `check_codes()` to ensure proper metadata.

3. **Geospatial Checks**  
   - Visualize points on a map with `plot_map_leaflet()`.  
   - Identify points on land using `check_onland()`.  
   - Optionally, use `positions_are_near_land()` and `which_basin()` for additional spatial QC.

4. **Depth Checks**  
   Run `check_depth()` to detect missing, invalid, or inconsistent depth values.

5. **Outlier Detection**  
   Use `check_outliers()` along with `scatterplot()` to identify statistical anomalies in environmental parameters and measurements.

6. **Station Matching**  
   Validate station information using `match_station()` to ensure correct station identifiers.

7. **Final Review & Visualization**  
   After all checks, plot your cleaned dataset on maps or scatterplots to confirm data quality.

Following this workflow ensures that SHARK occurrence data are consistent, valid, and ready for analysis.

---

## Citation

```{r, echo=FALSE}
# Print citation
citation("SHARK4R")
```
