---
title: "Retrieve Taxonomic Data From AlgaeBase"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retrieve Taxonomic Data From AlgaeBase}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---
  
## AlgaeBase
  
AlgaeBase is a comprehensive database containing information on a wide variety of algae species, including terrestrial, marine, and freshwater organisms, with an emphasis on marine botany. AlgaeBase is continually updated and funded by various phycological societies, with contributions from researchers and institutions worldwide. It can be accessed via a [web interface](https://www.algaebase.org/) or through the [API](https://www.algaebase.org/api/), as demonstrated in this tutorial using `SHARK4R`. Please note that the authors of `SHARK4R` are not affiliated with AlgaeBase.
  
## Getting Started
  
### Installation
  
You can install the latest version of `SHARK4R` from CRAN using:
```{r, eval=FALSE}
install.packages("SHARK4R")
```

Load the `SHARK4R` and `dplyr` libraries:
```{r, eval=FALSE}
library(SHARK4R)
library(dplyr)
```

```{r, include=FALSE}
suppressPackageStartupMessages({
  library(SHARK4R)
  library(dplyr)
})
```

### AlgaeBase API Key

AlgaeBase requires a subscription key to access its API. To obtain your own key, please visit the [API documentation](https://www.algaebase.org/api/). In the example below, the key is retrieved from an environment variable.

```{r}
# Retrieve the API key
algaebase_key <- Sys.getenv("ALGAEBASE_KEY")
```

## Match Genus Name

Taxonomic records can be retrieved for individual genera names using the `match_algaebase_genus()` function.

```{r}
# Match a genus name with AlgaeBase API
genus_records <- match_algaebase_genus(genus = "Gymnodinium",
                                       subscription_key = algaebase_key)

# Print the result
print(genus_records)
```

## Match Species Name

Taxonomic records can be retrieved for individual species names using the `match_algaebase_species()` function.

```{r}
# Match a species with AlgaeBase API
species_records <- match_algaebase_species(genus = "Tripos",
                                           species = "muelleri",
                                           subscription_key = algaebase_key)

# Print the result
print(species_records)
```

## Match Multiple Scientific Names

Multiple names can be matched with the `match_algaebase_taxa()` function. The scientific names need to be parsed into `genus` and `species` names before being passed to the API, which can be achieved by the `parse_scientific_names()` function.

```{r}
# Retrieve all phytoplankton data from April 2015
shark_data <- get_shark_data(fromYear = 2015, 
                             toYear = 2015,
                             months = 4, 
                             dataTypes = "Phytoplankton",
                             verbose = FALSE)

# Randomly select 10 rows from the shark_data dataframe
random_rows <- shark_data[sample(nrow(shark_data), 10), ]

# Parse scientific names into genus and species names
parsed_taxa <- parse_scientific_names(random_rows$scientific_name)

# Print the parsed data
print(parsed_taxa)

# Match the taxa with AlgaeBase
algaebase_match <- match_algaebase_taxa(genera = parsed_taxa$genus,
                                        species = parsed_taxa$species,
                                        subscription_key = algaebase_key,
                                        verbose = FALSE)

# Print the result
tibble(algaebase_match)
```

---

## Citation

```{r, echo=FALSE}
# Print citation
citation("SHARK4R")
```

```{r, echo=FALSE}
clean_shark4r_cache(0, clear_perm_cache = TRUE, verbose = FALSE)
```
