---
title: "Retrieve Taxonomic Data From Dyntaxa"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retrieve Taxonomic Data From Dyntaxa}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---
  
## Dyntaxa
  
Dyntaxa is a taxonomic database of Swedish organisms hosted at [SLU Artdatabanken](https://www.slu.se/artdatabanken/), providing information on their names and relationships. The database includes details such as the current classification, recommended names, and commonly used synonymous or misapplied names. Dyntaxa is continuously updated with new species for Sweden, new Swedish names, synonymous scientific names, and new data on relationships. The data in Dyntaxa serves as the foundation and framework for taxonomic information in [SHARK](https://shark.smhi.se/en/). It can be accessed via a [web interface](https://artfakta.se/) or through the [API](https://api-portal.artdatabanken.se/), as demonstrated in this tutorial using `SHARK4R`. Please note that the authors of `SHARK4R` are not affiliated with Dyntaxa.
  
## Getting Started
  
### Installation

You can install the latest version of `SHARK4R` from CRAN using:
```{r, eval=FALSE}
install.packages("SHARK4R")
```

Load the `SHARK4R` and `dplyr` libraries:
```{r, include=FALSE}
suppressPackageStartupMessages({
  library(SHARK4R)
  library(dplyr)
})
```

```{r, eval=FALSE}
library(SHARK4R)
library(dplyr)
```

### Retrieve Taxonomy Table from SHARK

Taxon and data tables can be retrieved with the same filtering options available in [SHARK](https://shark.smhi.se/en/). To see the available filtering options, please refer to `get_shark_options()` and the [Retrieve Data From SHARK](https://sharksmhi.github.io/SHARK4R/articles/retrieve_shark_data.html) tutorial.

```{r}
# Retrieve taxonomy reports for phytoplankton between 2019 and 2020
shark_taxon <- get_shark_data(tableView = "report_taxon",
                              fromYear = 2019,
                              toYear = 2020,
                              dataTypes = "Phytoplankton",
                              verbose = FALSE)

# Print data
print(shark_taxon)

# Retrieve all phytoplankton data from July 2015
shark_data <- get_shark_data(tableView = "sharkdata_phytoplankton",
                             fromYear = 2015, 
                             toYear = 2015,
                             months = 7,
                             dataTypes = "Phytoplankton",
                             verbose = FALSE)

# Print data
print(shark_data)
```

### Dyntaxa API Key

Dyntaxa requires a subscription key to access its API. To obtain your own key, sign up for the taxonomy product at the [SLU Swedish Species Information Centre's Developer Portal](https://api-portal.artdatabanken.se/). In the example below, the key is retrieved from an environment variable.

```{r}
# Retrieve the API key
dyntaxa_key <- Sys.getenv("DYNTAXA_KEY")
```

## Update SHARK Taxonomy Data

If the taxonomic data downloaded from SHARK are outdated, they can be updated to the latest Dyntaxa information using `SHARK4R`. Alternatively, data can be retrieved from WoRMS. For details, see the [WoRMS Tutorial](https://sharksmhi.github.io/SHARK4R/articles/retrieve_worms_data.html).

```{r}
# Update taxonomy information for the retrieved phytoplankton data
updated_taxonomy <- update_dyntaxa_taxonomy(
  dyntaxa_ids = shark_data$dyntaxa_id,
  subscription_key = dyntaxa_key,
  verbose = FALSE)

# Print the updated taxonomy data
print(updated_taxonomy)
```

## Match Taxon Names

```{r}
# Randomly select 10 phytoplankton taxa from shark_taxon
taxon_names <- sample(shark_taxon$scientific_name, size = 10)

# Match taxon_names with Dyntaxa API
matches <- match_dyntaxa_taxa(taxon_names = taxon_names, 
                              subscription_key = dyntaxa_key, 
                              multiple_options = FALSE,
                              verbose = FALSE)

# Print the result
tibble(matches)
```

## Retrieve Taxonomic information

Taxonomic records can be retrieved for individual taxa using the `get_dyntaxa_records()` function.

```{r}
# Get all Dyntaxa IDs
dyntaxa_id <- unique(matches$taxon_id)

# Remove potential NAs
dyntaxa_id <- dyntaxa_id[!is.na(dyntaxa_id)]

# Get Dyntaxa records
dyntaxa_records <- get_dyntaxa_records(taxon_ids = dyntaxa_id,
                                       subscription_key = dyntaxa_key)

# Print records
tibble(dyntaxa_records)
```

## Retrieve Parent IDs

All parent taxa above the Dyntaxa ID can be retrieved using the `get_dyntaxa_parent_ids()` function.

```{r}
# Get all parents
parents_id <- get_dyntaxa_parent_ids(taxon_ids = dyntaxa_id, 
                                     subscription_key = dyntaxa_key,
                                     verbose = FALSE)

# List the IDs
print(parents_id)
```

## Construct Complete Taxonomic Table

A comprehensive taxonomic table, including related taxa, can be created with the `construct_dyntaxa_table()` function. Use the `add_synonyms` parameter to include synonyms, and the `add_parents` and `add_descendants` parameters to include parent and descendant taxa, respectively. If Taxon IDs are missing from the DwC-A export downloaded through `get_dyntaxa_dwca()` (e.g. species complex and pseudotaxon), they can be matched using the `add_missing_taxa` argument. Additionally, complete hierarchy information can be added as a string of parent taxa separated by "-" using the `add_hierarchy` argument.

```{r}
# Retrieve complete taxonomic table (including parents and descendants)
taxonomy_table <- construct_dyntaxa_table(taxon_ids = dyntaxa_id, 
                                          subscription_key = dyntaxa_key, 
                                          shark_output = FALSE, 
                                          add_parents = TRUE,
                                          add_synonyms = TRUE, 
                                          add_descendants = TRUE,
                                          add_descendants_rank = "genus",
                                          add_missing_taxa = FALSE,
                                          add_hierarchy = FALSE,
                                          verbose = FALSE)

# Print the taxonomy table as a tibble
tibble(taxonomy_table)
```

---

## Citation

```{r, echo=FALSE}
# Print citation
citation("SHARK4R")
```
