---
title: "Retrieve Taxonomic Data From Dyntaxa"
output:
  rmarkdown::html_document:
    encoding: UTF-8
vignette: >
  %\VignetteIndexEntry{Retrieve Taxonomic Data From Dyntaxa}
  %\VignetteEngine{knitr::rmarkdown}
---

## Dyntaxa

Dyntaxa is a taxonomic database of Swedish organisms hosted at [SLU Artdatabanken](https://www.slu.se/artdatabanken/), providing information on their names and relationships. The database includes details such as the current classification, recommended names, and commonly used synonymous or misapplied names. Dyntaxa is continuously updated with new species for Sweden, new Swedish names, synonymous scientific names, and new data on relationships. The data in Dyntaxa serves as the foundation and framework for taxonomic information in [SHARK](https://shark.smhi.se/). It can be accessed via a [web interface](https://artfakta.se/) or through the [API](https://api-portal.artdatabanken.se/), as demonstrated in this tutorial using `SHARK4R`.

## Getting Started

#### Installation

You can install the package from GitHub using the `devtools` package:
```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("sharksmhi/SHARK4R",
                         dependencies = TRUE)
```

Load the `SHARK4R` and `tibble` libraries:
```{r}
library(SHARK4R)
library(tibble)
```

## Retrieve Taxonomic Data From SHARK

#### Retrieve Taxonony Table

Taxon tables can be retrieved with the same filtering options available in [SHARK](https://shark.smhi.se/). To see the available filtering options, please refer to [`get_shark_options`](../reference/get_shark_options.html) and the information below.

```{r}
# Retrieve chlorophyll data for April to June from 2019 to 2020
shark_taxon <- get_shark_data(tableView = "report_taxon",
                              fromYear = 2019,
                              toYear = 2020,
                              months = c(),
                              dataTypes = "Phytoplankton",
                              verbose = FALSE)

# Print data
print(shark_taxon)
```

## Retrieve Dyntaxa Data

#### Dyntaxa API Key

Dyntaxa requires a subscription key to access its API. To obtain your own key, sign up for the taxonomy product at the [SLU Swedish Species Information Centre´s Developer Portal](https://api-portal.artdatabanken.se/). In the example below, the key is retrieved from an environment variable.

```{r}
# Retrieve the API key
dyntaxa_key <- Sys.getenv("DYNTAXA_KEY")
```

#### Match taxon names

```{r}
# Randomly select 10 taxa from shark_taxon$scientific_name
taxon_names <- sample(shark_taxon$scientific_name, size = 10)

# Match taxon_names with Dyntaxa API
matches <- match_taxon_name(taxon_names, 
                            dyntaxa_key, 
                            multiple_options = FALSE,
                            verbose = FALSE)

# Print the result
tibble(matches)
```

#### Retrieve Taxonomic information

Taxonomic records can be retrieved for indivudual taxa using the [`get_dyntaxa_records`](../reference/get_dyntaxa_records.html) function.

```{r}
# Get all Dyntaxa IDs
dyntaxa_id <- unique(matches$taxon_id)

# Remove potential NAs
dyntaxa_id <- dyntaxa_id[!is.na(dyntaxa_id)]

# Get Dyntaxa records
dyntaxa_records <- get_dyntaxa_records(dyntaxa_id,
                                       dyntaxa_key)

# Print records
tibble(dyntaxa_records)
```

#### Retrieve Parent IDs

All parent taxa above the Dyntaxa ID can be retrieved using the [`get_dyntaxa_parent_ids`](../reference/get_dyntaxa_parent_ids.html) function.

```{r}
# Get all parents
parents_id <- get_dyntaxa_parent_ids(dyntaxa_id, 
                                     dyntaxa_key,
                                     verbose = FALSE)

# List the IDs
print(parents_id)
```

#### Construct Complete Taxonomic Table

A complete taxonomic table, including related taxa, can be constructed using the [`construct_dyntaxa_table`](../reference/construct_dyntaxa_table.html) function. Synonms can be included using the `recommended_only` parameter.

```{r}
# Retrieve complete taxonomic table (including parents and all descendants below genus level)
taxonomy_table <- construct_dyntaxa_table(parents_id, 
                                          dyntaxa_key, 
                                          shark_output = TRUE, 
                                          recommended_only = FALSE, 
                                          add_genus_children = TRUE,
                                          verbose = FALSE)

# Print the taxonomy table as a tibble
tibble(taxonomy_table)
```

## Citation

```{r, echo=FALSE}
# Print citation
citation("SHARK4R")
```
