---
title: "Retrieve Data From SHARK"
output:
  rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Retrieve Data From SHARK}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Getting Started

#### Installation

You can install the package from GitHub using the `devtools` package:
```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("sharksmhi/SHARK4R",
                         dependencies = TRUE)
```

Load the `SHARK4R` library:
```{r}
library(SHARK4R)
```

## Retrieve Data Using SHARK4R

#### Retrieve Data Table

Data can be retrieved with the same filtering options available in [SHARK](https://shark.smhi.se/). To see the available filtering options, please refer to [`get_shark_options`](../reference/get_shark_options.html) and the information below.

```{r}
# Retrieve chlorophyll data for April to June from 2019 to 2020
shark_data <- get_shark_table(fromYear = 2019, 
                              toYear = 2020,
                              months = c(4, 5, 6), 
                              dataTypes = c("Chlorophyll"),
                              limit = 200)

# Print data
print(shark_data)
```

### Get SHARK API Options

Filtering options, including data types, dataset names, stations, taxa, and more, can be retrieved using the [`get_shark_options`](../reference/get_shark_options.html) function.

```{r}
# Retrieve available search options
shark_options <- get_shark_options()

# List the names of the available options
names(shark_options)

# View available datatypes
dataTypes <- unlist(shark_options$dataTypes)
print(dataTypes)
```

## Citation

```{r, echo=FALSE}
# Print citation
citation("SHARK4R")
```
