---
title: "Retrieve HAB Data From IOC-UNESCO Databases"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retrieve HAB Data From IOC-UNESCO Databases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## IOC-UNESCO Databases

The Intergovernmental Oceanographic Commission (IOC) of UNESCO is the United Nations body dedicated to supporting global ocean science, services, and governance. The [IOC-UNESCO Taxonomic Reference List of Harmful Microalgae](https://www.marinespecies.org/hab/) focuses on species known to produce toxins or exhibit toxic effects, along with a few species suspected of toxin production. All toxic species in the list are verified with [WoRMS](https://www.marinespecies.org) to ensure accurate taxonomy.

The [IOC-UNESCO Toxins database](https://toxins.hais.ioc-unesco.org/) complements this list by providing detailed reference information about toxins, some of which are associated with harmful algal events.

The information from these databases can be access through APIs, as demonstrated in this tutorial using `SHARK4R`.

## Getting Started

#### Installation

You can install the latest version of `SHARK4R` from CRAN using:
```{r, eval=FALSE}
install.packages("SHARK4R")
```

Load the `SHARK4R` and `dplyr` libraries:
```{r, eval=FALSE}
library(SHARK4R)
library(dplyr)
```

```{r, include=FALSE}
suppressPackageStartupMessages({
  library(SHARK4R)
  library(dplyr)
})
```

## Retrieve HAB List From IOC Taxonomic List

The complete HAB list, including scientific names and AphiaIDs, can be downloaded from the [IOC-UNESCO Taxonomic Reference List of Harmful Microalgae](https://www.marinespecies.org/hab/). The output fields are customizable through function parameters—for example, setting `classification = FALSE` excludes higher taxonomic information from the results.

```{r}
# Retrieve complete HAB list
hab_list <- get_hab_list()

# Print result as tibble
tibble(hab_list)
```

## Retrieve HAB Toxins From IOC Toxins Database

The complete Toxin list can be downloaded from the [IOC-UNESCO Toxins database](https://toxins.hais.ioc-unesco.org/) using the `get_toxin_list()` function.

```{r}
# Retrieve complete toxin list
toxin_list <- get_toxin_list()

# Print result as tibble
tibble(toxin_list)
```

---

## Citation

```{r, echo=FALSE}
# Print citation
citation("SHARK4R")
```
